---
title: 'Data Cleaning: NYC School Perceptions'
author: "Mariby Cruz"
date: "2025-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## Introduction

What the project is about
What my goal is in this project.
Do student, teacher, and parent perceptions of NYC school quality appear to be related to demographic and academic success metrics?
Do students, teachers, and parents have similar perceptions of NYC school quality?

```{r}
#packages used
library(readr)
library(tidyverse)
library(tidyr)
library(purrr)
library(dplyr)
```

Do you have any thoughts on which variables will be necessary for your analysis?
Can any variables be excluded from the analysis?
Is there a variable you can use as a key to join the survey data to the combined dataset you've already prepared for analysis?

The data was retrieved from the 2011 NYC Schools Survey. Each line of data has information from each school in NYC, all school-levels and special instruction included. Each year, the survey is conducted and answered by parents, teachers, and students. It collects information on the perception of the quality of education, gives insight into the learning environments and diversification that goes beyond test scores. The questions in the survey assess: Safety & Respect, Communication, Engagement, and Academic Expectations.

```{r}
#importing data
combined <- read_csv("~/git/DataQuest_guided_projects/combined_schools.csv", show_col_types =  FALSE)

gened_df <- read_tsv("~/git/DataQuest_guided_projects/2011 data files online/masterfile11_gened_final.txt", show_col_types = FALSE)

district75_df <- read_tsv("~/git/DataQuest_guided_projects/2011 data files online/masterfile11_d75_final.txt", show_col_types = FALSE) 
```

### Simplifying the Dataframes

```{r}
#filtering for High school only, no mix of middle school & high school for example.
#interested in the perception of parents

highschool <- gened_df %>%
    filter(schooltype == "High School") %>% #narrowed observations from 1646 to 383
    select(-contains(c("_t", "_s"))) #dropped teacher and student responses
```

