---
title: 'Data Cleaning: NYC School Perceptions'
author: "Mariby Cruz"
date: "2025-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## Introduction

What the project is about
What my goal is in this project.
Do student, teacher, and parent perceptions of NYC school quality appear to be related to demographic and academic success metrics?
Do students, teachers, and parents have similar perceptions of NYC school quality?

```{r}
#packages used
library(readr)
library(tidyverse)
library(tidyr)
library(purrr)
library(dplyr)
```

Do you have any thoughts on which variables will be necessary for your analysis?
Can any variables be excluded from the analysis?
Is there a variable you can use as a key to join the survey data to the combined dataset you've already prepared for analysis?

The data was retrieved from the 2011 NYC Schools Survey. Each line of data has information from each school in NYC, all school-levels and special instruction included. Each year, the survey is conducted and answered by parents, teachers, and students. It collects information on the perception of the quality of education, gives insight into the learning environments and diversification that goes beyond test scores. The questions in the survey assess: Safety & Respect, Communication, Engagement, and Academic Expectations.

```{r}
#importing data
combined <- read_csv("~/git/DataQuest_guided_projects/nyc_school_survey/combined_schools.csv", show_col_types =  FALSE)

gened_df <- read_tsv("~/git/DataQuest_guided_projects/nyc_school_survey/2011 data files online/masterfile11_gened_final.txt", show_col_types = FALSE)

district75_df <- read_tsv("~/git/DataQuest_guided_projects/nyc_school_survey/2011 data files online/masterfile11_d75_final.txt", show_col_types = FALSE) 
```

### Simplifying the Dataframes

```{r}
#filtering for High school only, no mix of middle school & high school for example.
#interested in the perception of parents

highschool <- gened_df %>%
    filter(schooltype == "High School") %>% #narrowed observations from 1646 to 383
    select(1:32, -d75, -highschool) %>%#dropped teacher and student responses
    rename(DBN = dbn)

district75 <- district75_df %>% #narrowed observations from 1646 to 383
    select(1:32, -d75, -highschool) %>% #dropped teacher and student responses
    rename(DBN = dbn)
```

### Creating a Single Dataframe

```{r}
schools_combined <- rbind(highschool, district75) #since both df contained the same columns, rbind() used

combined_all <- combined %>%
    left_join(schools_combined, by = "DBN") #left joining to add data that matches DBN on the combined dataframe which has all testing scores

```

### Looking for Correlations and Examining Relationships

After creating a correlation matrix and filtering for the variables that have a correlation strength of -0.5 or +0.5, the data show 11 variables that influence the average SAT score. 
Among these are test variables like the average math, writing and reading scores - these have a strong positive relationship. 
Asian and white percentage also had a contribution to a high SAT score, but the relationship is not the strongest. 
It was interesting to see a strong negative relationship with the free or reduced lunch percent, meaning that it contributed to a low SAT score.
```{r}
#creating a correlation matrix to narrow in on relationships between variables
cor_mat <- combined_all %>%
  select(where(is.numeric)) %>% cor(use = "pairwise.complete.obs")
#visualizing the correlation matrix by turning the dataframe into a tibble
cor_tib <- cor_mat %>%
  as_tibble(rownames = "variable")
#followed by selecting those variables Plus Y-variable and filtering the cor strength.
sat_cors <- cor_tib %>%
  select(variable, avg_sat_score) %>%
  filter(avg_sat_score > 0.5 | avg_sat_score < -0.5) 

print(sat_cors)
```

There were 30 schools in the survey that did not have what Boro it was from, so these were removed from the plot. It did not change the relationship between average SAT score and schools with free and reduced lunch. 

This is an interesting observation, because I would have thought that having access to food would mean that a student would perform better. This may be misleading and might point to look into the types of testing resources available for students. 

Schools in the Bronx and Brooklyn had the highest percentage of free and reduced lunch but also the lowest average SAT scores. 
```{r}
frl_plot <- combined_all %>% 
  filter(!is.na(boro)) %>%
  ggplot(aes(x = frl_percent, y = avg_sat_score, color = boro)) +
  geom_point()

frl_plot
```

